---
title: "Lab 2 Solutions"
output: html_document
---





1. What happens when we mix types in R?
```{r}
age<-c(20, 15, 13, "ten") 
```

*You cannot mix numeric variables with text.  The entire vector is treated as text*

2. Can you do math on vectors in R?  Try the following. What happens and why?  [Uses code from above.]
```{r}
grades<-c(20, 15, 13, 3) 
percent<-(grades/25*100)
#age<-age+1   You can do math on vectors but in the above example, age is treated as a string vector.
```




3. Using R, write a set of commands that returns the sum, average, and standard deviation of a vector.

```{r}
ages<- c(19,19,12,23)
ages.sum<- sum(ages)
ages.mean<-mean(ages)
ages.sd<-sd(ages)
print(c(ages.sum, ages.mean, ages.sd))
```

4.  Using R, write a set of commands that swaps a set of variables such that A is equal to B and B is equal to A. 

```{r}
a<-4
b<-3
c<-a
a<-b
b<-c
print(c(a,b))
```

5. Describe the difference between a factor vector and a String vector in R.  

"Factor variables are categorical variables that can be either numeric or string variables. There are a number of advantages to converting categorical variables to factor variables. Perhaps the most important advantage is that they can be used in statistical modeling where they will be implemented correctly, i.e., they will then be assigned the correct number of degrees of freedom." [Source](http://www.ats.ucla.edu/stat/r/modules/factor_variables.htm)

6. Think back to our Titanic dataset.  Could we employ an R matrix for this data?  Why or why not? 

*No, because the data was of many different types.* 

7. In R, why is it relevant to set the working directory?  On the virtual machine, what directory corresponds with the course repository? 

*Setting the working directory provides a location of the files.  This is where R will look for a file. Files between the guest and the host will be shared at `/vagrant`.*

8. Specify the R code to do each of the following: (a.) a random number between 1 and 8.  (b.) a random integer between 1 and 10. (c.) a random set of 2 integers 1-10 that are not the same, (d.) a random number from a distribution with  a mean of 4 and a standard deviation of 2. 
**Hint: [This guide](http://blog.revolutionanalytics.com/2009/02/how-to-choose-a-random-number-in-r.html)  and [this guide](http://www.cookbook-r.com/Numbers/Generating_random_numbers/)can provide a nice overview.

```{r}
# A. A random number between 1 and 8. 
x1 <- runif(1, 1, 8)
x1 

# B. A random integer between 1 and 10.
 x2 <- sample(1:10, 1)
 x2
 
 # C. A random set of 2 integers 1-10 that are not the same
x3 <- sample(1:10, 2, replace=F)
x3

# D. A random number from a distribution with  a mean of 4 and a standard deviation of 2. 
x4 <- rnorm(1, mean=4, sd=2)
x4

```


9. What does it mean to sample random numbers "with replacement" vs. without replacement. 

*With replacement means random numbers can repeat. Without replacement means that the random numbers cannot repeat.*

12. Create a function in R that accepts 2 value (n, m) and returns an N x M matrix with random numbers with a mean of 20 and a standard deviation of 5.  

```{r}
n <- sample(1:9, 1)
m <- sample(1:9, 1)
test1 <- function(n,m) {matrix(rnorm(n*m, mean=20, sd=5), nrow=n, ncol=m)}
test1(n,m)
```
