---
title: "Technology Fundamentals for Business Analytics"
output:  html_document
---

## Overview 

We will provide here a number of different processes for *subsetting*, *linking*, and *aggregation* of data. There are a number of different reasons why you might want to do each of these processes.  

### Subsetting - Selecting Columns
We have already seen some level of selection in the "slicing" of vectors and matrix by location. This is an extremely important way  

```{r}
#Let's start by creating some data.
#This creates a vector with 10 numbers
v=(1:10)
v
#This creates a matrix 4,
m<- matrix(rnorm(40, mean=20, sd=5), nrow=10, ncol=4)
colnames(m)<-(c("a","b","c","d"))

#This creates a dataframe
df<- as.data.frame(m)

#We could also put together 
df2<- data.frame(v, m)


#Select Specific Columns
df[,1:2] #All rows, column 1-2
df[1:2]  #All rows, column 1-2
df[,c(1,3)] #We can also make the selection a vector.  This dynamically creates a vector.
columns.v<-c(1,3)
df[,columns.v] #We can also make the selection a vector.  This dynamically creates a vector.
df[,c("a","c")] #We can also do it by column name.
columns.b <- names(df) %in% c("a", "b", "c")
df[,columns.b]

#We can also drop specific columns
df[,c(-1,-4)] #drop column a and d
df[!columns.b]  #drop all but d

#There is also a specific subset function
subset(df, select=c(a, b) )

```

###Selecting Rows
At times we may want to split our sample for analysis.  We may just want to do it in order or we may want the process to be more random. 
 

```{r}
df[1:5,]  #The first 5 rows and all columns.
df[c(1,3,8),]  #Just a selection of a few rows
rows<-c(1,3,8)
df[rows,]
```
```{r}
#For a random sample, let's first create a random vector of the row numbers.
n<-nrow(df)
sample <- sample(10, 10, replace=F)
sample
```

```{r}
#notice how the array is being used to sample out the rows. 
df[sample[1:5],]
df[sample[6:10],]


#Remember we can select out the a variable of a dataframe using df$a
df[ which(df$a>=20.0), ] 
df[ which(df$a>=20 & df$b>=18.0), ]  # & is the and sign
df[ which(df$a>=20 | df$b>=18.0), ]  # | is the or sign

#We can also use the subset function to select out columes
subset(df, a >= 20 )
subset(df, a >= 20 & df$b>=18.0 )
subset(df, a >= 20 & df$b>=18.0 )

#We can also combine column and row selection as follows
subset(df, a >= 20 & df$b>=18.0, select=c(a, b) )
```


## Exercises
These toy examples are fine, but for this exercise we will use the Titaantic dataset. We want to split this out in a few different ways.  

1R. Subset only the *survived*, *sex*, and *age* columns and create a new dataframe with the results.

2R. Drop the *ticket* and the cabin and create a new dataframe with the results.

3R. Split the dataset into a *train* dataframe having the first 80% of the records and a *test* dataframe having the last 20%. 

4R. Randomely sample the dataset so that there are 2 samples.

5R. Create one dataframe with all of the *males* and a second with all of the *females*. 


### Getting CSV Files into RStudio

If your are having trouble getting file into RStudio, here are some options.

####OPTION 1
If sharing is working between the guest virual machine and the host, the easiest way is to sync your Git repository and then load the data from the shared drive. Look at the repository directory.
 
setwd('/vagrant/data')
list.files()
titantic <- read.csv('titantic_train.csv', header = TRUE )

###OPTION 2
You can import a dataframe from github directly using the gui with import dataset->from web url.  
[https://raw.githubusercontent.com/RPI-Analytics/MGMT6963-2015/master/data/titantic_train.csv](https://raw.githubusercontent.com/RPI-Analytics/MGMT6963-2015/master/data/titantic_train.csv)
Note. During the import process on the top left you can specify the dataframe name.  By default it will come through as titantic_train. You can create a new one called titantic using the code below.
`titantic<-titantic_train`


###OPTION 3
#R doesen't seem to have a real robust method of dealing with files and https, which I have seen cause some problems. 
Instead, you can `vagrant ssh` into the `/home/vagrant` directory of the virtual machine this is default. Then enter `wget https://raw.githubusercontent.com/RPI-Analytics/MGMT6963-2015/master/data/titantic_train.csv` from the terminal. This will download the file directly to the linux virtual machine. 
 
#setwd('/home/vagrant') You only need to do this if you changed the working directory to something else.
list.files()
titantic <- read.csv('titantic_train.csv', header = TRUE )
 



### Merging data
Merging datasets from different sources can be very common for a data scientist.  You might need to 


```{r}
key=(1:10)
 
#This creates a matrix 4,
#Here we are passing the row names and column names as a list. 
m<- matrix(rnorm(40, mean=20, sd=5), nrow=10, ncol=4, dimnames=list((1:10),c("a","b","c","d")))
m2<- matrix(rnorm(40, mean=1000, sd=5), nrow=10, ncol=4, dimnames=list((1:10),c("e","f","g","h")))
m3<- matrix(rnorm(40, mean=1000, sd=5), nrow=10, ncol=4, dimnames=list((1:10),c("e","f","g","h")))

#This creates a dataframe where the same key is across both
df<-  data.frame(key,m)
df2<- data.frame(key,m2)
key=(11:20)
df3<- data.frame(key,m3)


#slight alterhate syntax, using column bind first.
df4<-data.frame(cbind(key,m3))

#First let's merge rows for the same columns by combining df3 with rbind.  Key is that these have to have the same columns
rbind(df2,df3)

#let's say we have a dataframe that doesn't have the "e" column
df5<-df3[,-2] #Drop the e column
```
By then entering the `rbind(df2,df5)` command we will get the error: 
`Error in rbind(deparse.level, ...) : 
  numbers of columns of arguments do not match`

So instead we have to go ahead and readd the column as NA.

```{r}
df5$e=NA
df5
#Notice that it doesn't matter that columns are not in the same order.
rbind(df2,df5)
```

That merged some rows with the same structure.  But let's say we have to merge columns.  For that we need a common key.

```{r}
df6 <- merge(df,df2,by="key")
df6

#If we try to merge by a key were there is no match, it will result in 0 rows.
df7 <- merge(df,df3,by="key")
df7
```
But we can override this.  Let's say we want to include data from one or more dataframe even where there is no match.  This type of situation is called an *outer join* when working with SQL.

```{r}
df8 <- merge(df,df3,by="key",all.x=TRUE)   #x is df4 It will add NA for other (like left outer join)
df9 <- merge(df,df3,by="key", all.y=TRUE)  #y is df3 (like right outer join)
df10 <- merge(df,df3,by="key",all.x=TRUE, all.y=TRUE)  #y is df3 (like full outer join)
```
6R. You will find another dataset on github that contains additional rows of individuals [here](https://raw.githubusercontent.com/RPI-Analytics/MGMT6963-2015/gh-pages/assets/data/titantic_morerows.csv). (In reality they are duplicates but let's say we didn't know that). Provide the code to first import thist dataset and then combine it with the origional titantic dataframe into a new one called titantic_<yourlastname>. 

7R. You will find another dataset with additional variables that may help in the analysis [here](https://raw.githubusercontent.com/RPI-Analytics/MGMT6963-2015/gh-pages/assets/data/titantic_morecolumns.csv). Note. This is a little bit more tricky, as there currently isn't a key in the origional dataset. As a result, do the following substeps: (a). Provide code to verify the number of records in the origional and new dataset. (b). Generate a key for the origional dataset in the pattern of the new dataset. (c). Provide code to merge the two datasets titantic_cats_<yourlastname>.




### Aggregage Data  
Finally, we are going to aggregate data.  

```{r}

#This looks across all men and wormen on the titantic to average who survived.
#aggregate(titantic$survived, by=list(titantic$sex), FUN=mean, na.rm=FALSE)

#You can further recode a variable based on whether the 
#titantic$child <- ifelse(titantic$age > 18, c("adult"), c("child")) 
#aggregate(titantic$survived, by=list(titantic$sex, titantic$child), FUN=mean, na.rm=FALSE)

#This is a different recoding scheme where we are changing the value for the selected rows were the condition is met
#titantic$child[titantic$age > 18] <- "adult"
#titantic$child[(titantic$age > 2 & titantic$age <= 18)] <- "child"
#titantic$child[titantic$age <= 2] <- "infant"
#aggregate(titantic$survived, by=list(titantic$sex, titantic$child), FUN=mean, na.rm=FALSE)


```










